Assessing primitives performance\\ on multi-stage execution

### 2018-05-09 paper.tex
section comparison: about optimization potential

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (-) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \todo{}
	discuss inlining C comp cross file and InterpreterProxy indirection
	Smalltalk we control indirectly.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

>>>>

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (-) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} 
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.

	\todo{}
	Smalltalk we control indirectly.\sk{?}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}
